Precise, Second-Order Verification

### 2014-11-13 slides/nus-talk.tex
NUS slides

\lstset{language=C,
 basicstyle=\tiny
 }

@@

@@

\iffalse
\begin{frame}{Outline}
  \tableofcontents
\end{frame}
\fi

\section{Motivation}

@@

\begin{frame}[fragile]{Heap}
 \centering
\begin{tikzpicture}
  %\node at (-.75,.25) {$h:$};

@@

  \path[dashed,->] (x) edge (n2)%
		   (y) edge (n5)%
		   (z) edge (n1)%
		   (null) edge (n6);
 \end{tikzpicture}
\end{frame}

\begin{frame}[fragile]{Heap Transformers}
 \centering
 \begin{tikzpicture}
  \node (n1) {$\bullet$};
  \node [right of=n1] (n2) {$\bullet$};
  \node (n3) [right of=n2] {$\bullet$};

@@

  \path[draw, ->] (m1) edge (m2) (m2) edge (m3);
  \path[dashed,->] (x) edge (m2) (y) edge (m1) (null) edge (m3);
 \end{tikzpicture}
 }

 \vfill
 $\hlookup(h, x, y)$ \\
 $x = y{\rightarrow}next;$
\end{frame}

>>>>

\lstset{language=C,
 basicstyle=\tiny,
 mathescape=true
 }

@@

\usepackage{amsmath}
\usepackage{xspace}

@@

\begin{frame}{Outline}
  \tableofcontents
\end{frame}

\section{Program Synthesis and Second-Order Logic}

\section{Termination}

@@

\section{Heap}

\begin{frame}[fragile]{Heaps of Singly Linked Lists}
We would like to use second-order logic to analyse programs using the heap.  To do so, we need a
heap theory with a decision procedure in NP that we can combine with arithmetic.

We will restrict ourselves to singly linked lists with lengths, but no data.

Lists can have (unspecified) cycles and sharing, for example:

\begin{center}
\begin{tikzpicture}
  %\node at (-.75,.25) {$h:$};

@@

  \path[dashed,->] (x) edge (n2)%
		   (y) edge (n5)%
		   (z) edge (n1)%
		   (null) edge (n6);
 \end{tikzpicture}
\end{center}
\end{frame}

\begin{frame}[fragile]{Heap Transformers}
\only<1>{
x = y$\rightarrow$next

\begin{center}
 \begin{tikzpicture}
  \node (n1) {$\bullet$};
  \node [right of=n1] (n2) {$\bullet$};
  \node (n3) [right of=n2] {$\bullet$};

@@

  \path[draw, ->] (m1) edge (m2) (m2) edge (m3);
  \path[dashed,->] (x) edge (m2) (y) edge (m1) (null) edge (m3);
\end{tikzpicture}
\end{center}
}

\only<2>{
x = new()

\begin{center}
\begin{tikzpicture}[shorten >=1pt,auto]
  \node (n1) {$\bullet$};
  \node (n2) [right of=n1] {$\bullet$};
  \node (n3) [right of=n2] {$\bullet$};

  \node (x) [below of=n1] {$x$};
  \node (null) [below of=n3] {\bf null};

  \path[draw, ->] (n1) edge (n2)
                  (n2) edge (n3);
  \draw[dashed,->] (x) to (n1);
  \draw[dashed,->] (null) to (n3);

  \node [below right of=n3] (mid) {$\Rightarrow$};

  \node (m1) [above right of=mid]{$\bullet$};
  \node (m2) [right of=m1] {$\bullet$};
  \node (m3) [right of=m2] {$\bullet$};
  \node (m4) [below of=m2] {$\bullet$};

  \node (x2) [below =1em of m4] {$x$};
  \node (null) [right of=x2] {\bf null};
each of these 
  \path[draw, ->] (m1) edge (m2) (m2) edge (m3) (m4) edge (m3);
  \draw[dashed,->] (x2) to (m4);
  \draw[dashed,->] (null) to (m3);
 \end{tikzpicture}
 \end{center}
 }

 \only<3>{
x = y

\begin{center} 
 \begin{tikzpicture}[shorten >=1pt,auto]
  \node (n1) {$\bullet$};
  \node (n2) [right of=n1] {$\bullet$};
  \node (n3) [right of=n2] {$\bullet$};
  \node (n4) [below=2em of n2] {$\bullet$};

  \node (y) [below=1em of n4] {$y$};
  \node (x) [left of=y] {$x$};
  \node (null) [right of=y] {\bf null};

  \path[draw, ->] (n1) edge (n2)
                  (n2) edge (n3)
                  (n4) edge (n3);
  \path[dashed,->] (x) edge (n1)
                   (y) edge (n4)
                   (null) edge (n3);

  \node [below right of=n3] (mid) {$\Rightarrow$};

  \node (m1) [above right of=mid]{$\bullet$};
  \node (m2) [right of=m1] {$\bullet$};
  \node (m3) [right of=m2] {$\bullet$};
  \node (m4) [below=2em of m2] {$\bullet$};

  \node (y2) [below=1em of m4] {$y$};
  \node (x2) [left of=y2] {$x$};
  \node (null) [right of=y2] {\bf null};

  \path[draw, ->] (m1) edge (m2)
                  (m2) edge (m3)
                  (m4) edge (m3);
  \path[dashed,->] (x2) edge (m4)
                   (y2) edge (m4)
                   (null) edge (m3);
 \end{tikzpicture}
 \end{center}
 }

\only<4>{
x$\rightarrow$next = y

\begin{center}
 \begin{tikzpicture}[shorten >=1pt,auto]
  \node (n1) {$\bullet$};
  \node (n2) [right of=n1] {$\bullet$};
  \node (n3) [right of=n2] {$\bullet$};
  \node (n4) [below=2em of n2] {$\bullet$};

  \node (y) [below=1em of n4] {$y$};
  \node (x) [left of=y] {$x$};
  \node (null) [right of=y] {\bf null};

  \path[draw, ->] (n1) edge (n2)
                  (n2) edge (n3)
                  (n4) edge (n3);
  \path[dashed,->] (x) edge (n1)
                   (y) edge (n4)
                   (null) edge (n3);

  \node [below right of=n3] (mid) {$\Rightarrow$};

  \node (m1) [above right of=mid]{$\bullet$};
  \node (m2) [right of=m1] {$\bullet$};
  \node (m3) [right of=m2] {$\bullet$};
  \node (m4) [below=2em of m2] {$\bullet$};

  \node (y2) [below=1em of m4] {$y$};
  \node (x2) [left of=y2] {$x$};
  \node (null) [right of=y2] {\bf null};

  \path[draw, ->] (m1) edge (m4)
                  (m2) edge (m3)
                  (m4) edge (m3);
  \path[dashed,->] (x2) edge (m1)
                   (y2) edge (m4)
                   (null) edge (m3);
 \end{tikzpicture}
\end{center}
}

\begin{frame}[fragile]{Heap Observations}

\begin{center}
\begin{minipage}[c]{.4\textwidth}
\begin{tikzpicture}[>=stealth',shorten >=1pt,auto,scale=.8]
  %\node at (-.75,.25) {$h:$};

  \node (n1) {$\bullet$};
  \node (n2) [below of=n1] {$\bullet$};
  \node (n3) [right of=n1] {$\bullet$};
  \node (n4) [right of=n3] {$\bullet$};
  \node (n5) [right of=n4,above of=n4] {$\bullet$};
  \node (n6) [below of=n4] {$\bullet$};
  \node (x) [below of=n2] {$x$};
  \node (y) [right of=n5] {$y$};
  \node (z) [above of=n1] {$z$};
  \node (null) [below of=n6] {\bf null};

  \path[dashed,->] (x) edge (n2)%
		   (y) edge (n5)%
		   (z) edge (n1)%
		   (null) edge (n6);
 \end{tikzpicture}
 \end{minipage}
 \begin{minipage}[c]{.4\textwidth}
 \vfill
 \begin{align*}
 \hpathlen(x, y) & =  3 \\
 \hispath(z, y) & = \text{true} \\
 \hispath(x, z) & = \text{false} \\
 \halias(x, z) & = \text{false} \\
 \hisnull(x) & = \text{false} \\
 \hcircular(y) & = \text{true}
 \end{align*}
 \vfill
 \end{minipage}
\end{center}
\end{frame}

\begin{frame}[fragile]{The SLH Theory}
 SLH is our theory of singly linked lists.  SLH formulae can contain arbitrary boolean combinations of the
 predicates from the previous slides.

 For example:

 \[
  h' = \halloc(h, x) \wedge \hisnull(h', x)
 \]

\begin{frame}{\shakira}
 We built a decision procedure for SLH named \shakira, which checks that heaps don't lie.

 \shakira works by exploiting a small model property of SLH to build a SAT formula that is equisatisfiabile with the SLH formula.
\end{frame}

\begin{frame}[fragile]{Edge Subdivision}
\begin{center}
 \begin{tikzpicture}
  \node (h) {$H:$};

  \node (n1) [below right =.1em and .1em of h] {$\bullet$};
  \node (space) [right of=n1] {};
  \node (n2) [right of=space] {$\bullet$};

  \node (x) [below of=n1] {$x$};
  \node (null) [below of=n2] {\bf null};

  \path[draw, ->] (n1) to node[above] {3} (n2);
  \path[dashed,->] (x) edge (n1)
                   (null) edge (n2);

  \node (h2) [right=9em of h] {$H':$};

  \node (m1)[below right =.1em and .1em of h2] {$\bullet$};
  \node (m2) [right of=m1] {$\bullet$};
  \node (m3) [right of=m2] {$\bullet$};

  \node (x2) [below of=m1] {$x$};
  \node (null2) [below of=m3] {\bf null};

  \path[draw, ->] (m1) edge node[above] {1} (m2)
                  (m2) edge node [above] {2} (m3);
  \path[dashed,->] (x2) edge (m1)
		   (null2) edge (m3);
 \end{tikzpicture}
\end{center}

 \vspace{1em}
 $H'$ can be generated from $H$ by subdividing an edge.  We say $H'$ is a \emph{subdivision} of $H$.
\end{frame}

\begin{frame}[fragile]{Heap Homeomorphism}

 \begin{center}
 \resizebox{\textwidth}{!}{
 \begin{tikzpicture}
  \node (h) {$H:$};

  \node (x) [below right =.1em and .1em of h] {$x$};
  \node (y) [below of=x] {$y$};

  \node (n1) [right of=x] {$\bullet$};
  \node (n2) [below of=n1] {$\bullet$};

  \path (n1) to node (mid) {} (n2);

  \node (n3) [right of=mid] {$\bullet$};
  \node (n4) [right of=n3] {$\bullet$};

  \node (n5) [right of=n4] {$\bullet$};
  \node (space2) [right of=n5] {};
  \node (n6) [right of=space2] {$\bullet$};

  \path[->,draw] (n1) edge node [above] {1} (n3)
                 (n2) edge node [below] {2} (n3)
                 (n3) edge node [above] {4} (n4)
                 (n4) edge node [above] {2} (n5)
                 (n5) edge[bend right=70] node [below] {6} (n6)
                 (n6) edge[bend right=70] node [above] {3} (n5);

  \path[->,dashed] (x) edge (n1)
                   (y) edge (n2);

  \node (h2) [right=20em of h] {$H':$};

  \node (x2) [below right =.1em and .1em of h2] {$x$};
  \node (y2) [below of=x2] {$y$};

  \node (m1) [right of=x2] {$\bullet$};
  \node (m2) [below of=m1] {$\bullet$};

  \path (m1) to node (mid2) {} (m2);

  \node (m3) [right of=mid2] {$\bullet$};
  \node (m4) [right of=m3] {};

  \node (m5) [right of=m4] {$\bullet$};
  \node (space2) [right of=m5] {};
  \node (m6) [right of=space2] {$\bullet$};

  \node (m7) [above of=space2] {$\bullet$};

  \path[->,draw] (m1) edge node [above] {1} (m3)
                 (m2) edge node [below] {2} (m3)
                 (m3) edge node [above] {6} (m5)
                 (m5) edge[bend right=70] node [below] {6} (m6)
                 (m6) edge[bend right=35] node [above] {1} (m7)
                 (m7) edge[bend right=35] node [above] {2} (m5);

  \path[->,dashed] (x2) edge (m1)
                   (y2) edge (m2);

  \node (s) [below right=7em and 8em of h] {$S:$};

  \node (x) [below right =.1em and .1em of s] {$x$};
  \node (y) [below of=x] {$y$};

  \node (n1) [right of=x] {$\bullet$};
  \node (n2) [below of=n1] {$\bullet$};

  \path (n1) to node (mid) {} (n2);

  \node (n3) [right of=mid] {$\bullet$};
  \node (n4) [right of=n3] {$\bullet$};

  \node (n5) [right of=n4] {$\bullet$};
  \node (space2) [right of=n5] {};
  \node (n6) [right of=space2] {$\bullet$};

  \node (n7) [above of=space2] {$\bullet$};

  \path[->,draw] (n1) edge node [above] {1} (n3)
                 (n2) edge node [below] {2} (n3)
                 (n3) edge node [above] {4} (n4)
                 (n4) edge node [above] {2} (n5)
                 (n5) edge[bend right=70] node [below] {6} (n6)
                 (n6) edge[bend right=35] node [above] {1} (n7)
                 (n7) edge[bend right=35] node [above] {2} (n5);

  \path[->,dashed] (x) edge (n1)
                   (y) edge (n2);
% 
%   \node (x) [below of=n1] {$x$};
%   \node (null) [below of=n2] {\bf null};
% 
%   \path[draw, ->] (n1) to node[above] {3} (n2);
%   \path[dashed,->] (x) edge (n1) 
%                    (null) edge (n2);
% 
%   \node (h2) [right=9em of h] {$H':$};
% 
%   \node (m1)[below right =.1em and .1em of h2] {$\bullet$};
%   \node (m2) [right of=m1] {$\bullet$};
%   \node (m3) [right of=m2] {$\bullet$};
% 
%   \node (x2) [below of=m1] {$x$};
%   \node (null2) [below of=m3] {\bf null};
% 
%   \path[draw, ->] (m1) edge node[above] {1} (m2)
%                   (m2) edge node [above] {2} (m3);
%   \path[dashed,->] (x2) edge (m1)
%                    (null2) edge (m3);
 \end{tikzpicture}
 }
 \end{center}

 Both $H$ and $H'$ can be subdivided to produce $S$.  We then say that $H$ and $H'$ are \emph{homeomorphic}.
\end{frame}

\begin{frame}{SLH Has Small Models}
 \begin{theorem}[Homeomorphism Preserves Models]
  If two heaps are homeomorphic, they satisfy exactly the same SLH formulae.
 \end{theorem}

 \begin{theorem}[Homeomorphisms Can Be Small]
  Every homeomorphism class has a least element, which has no more than $2P$ nodes ($P$ is the number of program variables).
 \end{theorem}

 From this it follows that SLH has small models -- we only need to check models up to size $2P$.
\end{frame}
